---
title: "Homework: Building and Evaluating a Classification Model"
author: "Patrick Torralba"
date: "04/07/2026"
output: 
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, message=FALSE, warning=FALSE}
library(tidymodels)
library(tidyverse)
library(modeldata)
library(rpart.plot)
library(psych)

set.seed(25)
```

# Overview

In this assignment, you will build a **classification workflow from scratch** using a new dataset. AS with each homeowrk,w e are graduall stepping back the scaffolding as you get stronger as individual programmers. In this homework, you will need to make some modeling choices on your own and explain them clearly.

Your work should show that you can:

- identify and describe a classification problem
- split data into training and testing sets
- build a classification model using `tidymodels`
- generate both class predictions and probability predictions
- evaluate a classifier with a confusion matrix, metrics, and an ROC curve
- create your own `ggplot2` visualizations of model performance
- compare two different classification models
- reflect on your modeling decisions

You should write code in the code chunks and write your written responses in the spaces provided.

Before you begin the assignment:

1. Knit this document and make sure it runs.
2. Update the author name in the YAML.
3. Commit and push your work regularly as you go.

## Submission instructions

- Complete this file by adding code and short written responses where requested.
- Knit to a GitHub (`md`) document.
- Submit both the `.Rmd` file and the knitted `.md` file by committing to your repository. 

# Dataset

For this homework, use the **`credit_data`** dataset from the `modeldata` package.

The outcome you will model is whether an applicant is a **good** or **bad** credit risk.

## Load and inspect the data

Use the chunk below to load the data and inspect its structure.

```{r load-data}
data(credit_data)
glimpse(credit_data)
```

## Initial orientation

In a few sentences, describe what you notice about the dataset.

Prompt yourself with these questions:

- What is the name of the outcome variable?
- Is the outcome categorical or numeric?
- Do you see a mix of numeric and categorical predictors?
- Does anything stand out as potentially useful for prediction?

**Response:**

The dataset has 4454 rows and 14 columns. Most of the variables are numerical but a handful such as status, home, marital, records, and job are categorical predictors. The outcome is categorical (e.g good / bad). Variables that are useful for prediction include job, expenses, income, assests, debt, amount, and price.
---

# Part 1: Understanding the Classification Problem

## 1.1 Identify the outcome

Write code that helps you explore the outcome variable.

Your code should:

- identify the outcome variable
- show its values
- help you determine whether this is a classification problem

```{r outcome-exploration}

summary(credit_data$Status)


ggplot(credit_data, aes(x = Status, fill = Job)) +
  geom_bar() +
  facet_wrap(~ Job) +
  labs(x = "Credit Status",
       y = "Count",
       fill = "Job Type",
       title = "Credit Status by Job Type")

```

After running your code, answer the following:

1. What is the name of the outcome variable?
2. What are the possible classes?
3. Why is this a classification problem rather than a regression problem?

**Response:**

The name of the outcome variable is Status. The possible classes are either good or bad. This is a classification problem because Status is a binary variable - it can only be good or bad - and we are interested in predicting the probability of an individual either being classified into good or bad.

## 1.2 Examine class balance

Create:

- a table of the outcome variable
- a bar plot of the outcome variable

```{r}
#| label: tbl-status
#| tbl-cap: "Status Table"
#| tbl-cap-location: top

credit_data %>% 
  count(Status) %>% 
  mutate(Percent = (n / sum(n)) *100)

```



```{r class-balance}

ggplot(credit_data, aes(x = Status, fill = Status)) +
  geom_bar()


```

Then answer:

- Is the dataset balanced across the two classes, or is one class more common?
- Why might class balance matter when evaluating a classifier?
- Why might accuracy be misleading if one class is much more common?

**Response:**

The dataset is not balanced. There are more people with good credit than bad credit. Class balance matters when evaluating because this directly impacts accuracy. To be specific, since the classes are imbalanced, a logistic regression model only learns to predict the majority class (good), and not the minority class (bad). Accuracy may be misleading because we can only be accurate pertaining to the good credit since we have alot of data on it. But, we cannot be accurate on the bad credit since there isn't alot of data to go off of.

---

# Part 2: Prepare the Data

## 2.1 Create a train/test split

Split the data into **training** and **testing** sets.

Expectations:

- Use an 80/20 split.
- Use `set.seed(123)` if needed again.
- Use **stratification on the outcome variable** so that class proportions are preserved.

```{r split-data}

set.seed(123)

data_split <- initial_split(credit_data,
                    prop = 0.80,
                    strata = Status)

train_data <- training(data_split)

test_data <- testing(data_split)

```

Then answer:

- Why do we split data into training and testing sets?
- Why is stratifying on the outcome useful in this case?

**Response:**

We split data into training and testing datasets because we are trying to train the model by using the training data. And then we assess the model by verifying it with the testing data. Stratifying is useful because it ensures that we have an equal distribution of Status across both datasets.

## 2.2 Choose predictors

You will begin by building a **logistic regression model** using **at least 3 predictors** of your choice.

Before you write the model, briefly explain:

- which predictors you plan to use
- why you chose them
- whether there are any variables you intentionally decided not to use

You do **not** need to choose the “best” predictors, but your choices should be thoughtful and clearly explained.

**Response:**

I choose the following predictor variables: Income, Debt, and Job. I selected these variables because I believe they can strongly influence an individual's credit status. For example, an individual's job and income direclty impacts their spending habits, and thus to a degree, their credit status. I chose not to use variables like age, marital, and seniority because I felt that they weren't as powerful predictors as the the aforementioend selected variables. 


---

# Part 3: Build a Logistic Regression Classifier

## 3.1 Specify the model

Create a **logistic regression specification** using `tidymodels`.

```{r logistic-spec}

log_spec <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification") 
  

```

## 3.2 Build a workflow

Create a workflow that:

- adds your model
- adds a formula with the outcome and your chosen predictors

```{r logistic-workflow}

log_workflow <- workflow() %>% 
  add_model(log_spec) %>% 
  add_formula(Status ~ Income + Job + Debt)

log_workflow


```

## 3.3 Fit the model

Fit your workflow on the training data.

```{r logistic-fit}

log_fit <- log_workflow %>% 
  fit(data = credit_data)


log_fit
```

Then answer:

- Why are we fitting on the training set and not the full dataset?
- What is the benefit of keeping the model and formula inside a workflow?

**Response:**

We are fitting data on the training set only because we are training our model to recognize patterns. If we used the entire dataset then we'd have no test data to verify our model since the model has access to the entire dataset. The benefit of keeping the model and formula inside a workflow is that we can apply this model to different datasets (e.g another credit_data dataset).

---

# Part 4: Generate Predictions

Use your fitted logistic regression model to generate predictions on the **test** set.

You should generate:

- predicted probabilities
- predicted classes

Then combine those predictions with the original test data.

```{r logistic-predictions}

log_credit_status_preds <- predict(log_fit, new_data = test_data)

log_credit_prob_preds <-predict(log_fit, new_data = test_data, type = "prob")

log_results <- test_data %>% 
  select(Status) %>% 
  bind_cols(log_credit_status_preds, log_credit_prob_preds)


log_results
```

After generating predictions, answer:

- What is the difference between a predicted probability and a predicted class?
- Why is it useful to keep both?

**Response:**

A predicted probability gives the decimal value of the likelihood an individual is to either have good credit or bad credit. The predicted class is an absolute that is determined by the good / bad decimal value. For example, if the predicted good value is higher than the predicted bad value, then the predicted status is good.

It is useful to keep both because the probabilities allow you to see how the distribution whereas the class is needed for an absolute interpretation.

---

# Part 5: Evaluate the Logistic Regression Model

## 5.1 Confusion matrix

Create a confusion matrix for your logistic regression predictions on the test set.

```{r logistic-confusion-matrix}

log_results %>% 
  conf_mat(truth = Status, estimate = .pred_class)


```

Then answer:

- What do the rows represent?
- What do the columns represent?
- Where do correct predictions appear?
- What kinds of mistakes does your model make?

**Response:**

The rows represent the actual values whereas the columns represent the predicted values. The correct predictions are located on the symmetrical intersections (e.g bad x bad and good x good). The model can make false negatives and false positive mistakes.

## 5.2 Performance metrics

Use `metrics()` to calculate performance measures for your logistic regression model.

```{r logistic-metrics}
log_results %>% 
  metrics(truth = Status, estimate = .pred_class)
```

Then answer:

- What is the accuracy?
- What other metrics are shown?
- Is accuracy alone enough to judge performance here? Why or why not?

**Response:**

Accuracy is how close the model was from estimating the actual values. The other metric is kap. Accuracy alone is not enough to judge performance considering our model is only 77% accurate. We need to evaluate the ROC AUC as well.

---

# Part 6: Visualize Model Performance with ggplot2

In class, we discussed that a confusion matrix can be treated like a dataset and visualized using the grammar of graphics. In this section, build your own performance visualizations rather than relying only on built-in plots.

## 6.1 Confusion matrix heatmap using counts

Create a `ggplot2` heatmap of the confusion matrix using **counts**.

Expectations:

- convert the confusion matrix to a tibble
- map predicted class to one axis and true class to the other
- use `geom_tile()`
- display the count in each cell with `geom_text()`

```{r confusion-matrix-count-plot}

cm <- log_results %>% 
  conf_mat(truth = Status, estimate = .pred_class) 


cm_df <- as.data.frame(cm$table)

ggplot(cm_df, aes(x = Prediction, y = Truth, fill = Freq)) +
  geom_tile(alpha = 0.5) +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  labs(x = "Predicted Classes",
       y = "True Class",
       fill = "Count",
       title = "Confusion Matrix")

```

Then answer:

- What variable did you map to `fill`?
- Why is `geom_tile()` appropriate here?
- What patterns are easiest to notice in this version?

**Response:**

I mapped the the Freq variable to fill so it would show the counts in my geom_tile() plot. Geom_tile is appropriate because we are showing a 2x2 table that shows a visualization of how accurate our model is. The patterns that are super easy to notice are the counts and colors. For example, it is very easy to see that are good x good cell has 587 counts and is light blue.

## 6.2 Confusion matrix heatmap using proportions

Now create a second version of the confusion matrix plot using **proportions within each true class** instead of raw counts.

This means you should:

- group by the true class
- calculate a proportion for each cell within that class
- plot those proportions in a heatmap
- label cells with rounded proportions

```{r confusion-matrix-proportion-plot}

conf_prop_status_conf_mat <- conf_mat(log_results, truth = Status, estimate = .pred_class)


conf_prop_status_df <- as.data.frame(conf_prop_status_conf_mat$table) %>% 
  group_by(Truth) %>% 
  mutate(prop = (Freq / sum(Freq)))
  

ggplot(conf_prop_status_df, aes(x = Prediction, y = Truth, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(prop, 2)), color = "black") +
  labs(
    x = "Predicted Classes",
    y = "True Class",
    fill = "Proportion",
    title = "Confusion Matrix (proportions)"
    
  )


```

Then answer:

- Why might proportions be more useful than counts?
- What would strong performance look like in this plot?
- Do the proportions suggest similar performance across both classes?

**Response:**

Proportions might be more useful than counts because we can compare the false positives and false negatives against the true values. A strong performance would look like having really high proportion values and very light blue colors in the bad x bad and the good x good cells. The proportions do not suggest similar performances across both classes. This matrix indicates that the model is really effective at predicting "Good" credit status but ineffective at predicting "Bad" credit status.


## 6.3 Probability distribution plot

Create a plot showing predicted probabilities by actual class.

A histogram is one good option, but you may choose another `ggplot2` visualization if you prefer.

Your goal is to show how well the model separates the classes based on predicted probability.

```{r probability-plot}

ggplot(log_results, aes(x = .pred_good, fill = Status)) +
  geom_density() +
  labs(x = "Predicted Probability of 'Good'",
       y = "Density",
       fill = "True Status",
       title = "Predicted Probablity Distribution by True Status")


```

Then answer:

- How well does your model separate the two classes?
- Where do you see overlap?
- What does overlap suggest about misclassification?

**Response:**

It appears that the model is unable to separate the two classes well. The peaks for good and bad are somewhat close to each other, and thus, this represents that the model has difficulty differentiating the two classes. I see the overlap occurring at the 70% probability for "Good". This overlap represents a false negative and a false positive. A false negative for every "bad" individual that falls to the right of 0.5, and a false positive for every "good" falling left of 0.5. 

---

# Part 7: ROC Curve and AUC

Use your logistic regression predictions to create an ROC curve and calculate AUC.

Your code should:

- create an ROC curve using the probability for the positive class
- plot the ROC curve
- calculate AUC

```{r roc-and-auc}
# Write code here
```

Then answer:

- What does the ROC curve show that the confusion matrix does not?
- What is your AUC value?
- How would you interpret that AUC value in plain language?

**Response:**

_Type your response here._

---

# Part 8: Build a Second Model

In this section, you will build a **decision tree classifier** using the **same predictors** you used in logistic regression. This will allow a fairer comparison between the two models.

## 8.1 Specify and fit a decision tree

Create a decision tree model using `decision_tree()` with the `rpart` engine.

Then build a workflow and fit it on the training data.

```{r tree-fit}
# Write code here
```

Then answer:

- How is a decision tree conceptually different from logistic regression?
- What kinds of patterns might a tree capture more easily?

**Response:**

_Type your response here._

## 8.2 Visualize the tree

Visualize your fitted tree using `rpart.plot()`.

```{r tree-plot}
# Write code here
```

Then answer:

- What variable appears in the first split?
- What do the terminal nodes represent?
- Why might a tree be easier to interpret than logistic regression?

**Response:**

_Type your response here._

## 8.3 Generate predictions for the tree

Create both probability and class predictions for the tree on the test set, then combine them with the test data.

```{r tree-predictions}
# Write code here
```

## 8.4 Evaluate the tree

Evaluate the tree using:

- a confusion matrix
- `metrics()`

```{r tree-evaluation}
# Write code here
```

Then answer:

- How does the tree perform compared to logistic regression?
- Which model has stronger performance?
- Which model is easier to interpret?
- If you had to choose one of the two models, which would you choose and why?

**Response:**

_Type your response here._

---

# Part 9: Reflection

Write full-sentence responses to the following:

1. What was the most challenging part of this assignment?
2. What modeling decisions did you make that may have affected performance?
3. If you had more time, what would you try next?
4. How could this workflow apply to your own research, professional work, or interests?

**Response:**

_Type your response here._

---

# Submission

Submit:

1. your completed `.Rmd` file
2. your knitted HTML file

Make sure that:

- all code runs from top to bottom
- all plots appear in the knitted output
- your written responses are included
- your file is clearly named
